---
title: ""
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    page-layout: full
    embed-resources: true
    css: report/sample-report.css
    include-after-body: report/escape-report.html
execute:
  echo: false
  warning: false
  message: false
params:
  results_dir: "../../results"
  samples: ""
  run_summary_tsv: "run_summary.tsv"
  review_tsv: "samples_requiring_review.tsv"
  coverage_threshold: 50
  coverage_breadth_threshold: 0.95
  max_n_fraction: 0.01
---

```{r setup}
library(readr)
library(dplyr)
library(ggplot2)
library(htmltools)
library(knitr)
library(gt)

coverage_threshold <- as.numeric(params$coverage_threshold)
coverage_breadth_threshold <- as.numeric(params$coverage_breadth_threshold)
max_n_fraction <- as.numeric(params$max_n_fraction)

sample_ids <- trimws(strsplit(params$samples, ",", fixed = TRUE)[[1]])
sample_ids <- sample_ids[nzchar(sample_ids)]

if (length(sample_ids) == 0) {
  stop("No sample IDs were supplied through the 'samples' parameter.")
}

sample_summary_paths <- file.path(
  params$results_dir,
  sample_ids,
  "summary",
  paste0(sample_ids, ".sample_summary.tsv")
)

missing_paths <- sample_summary_paths[!file.exists(sample_summary_paths)]
if (length(missing_paths) > 0) {
  stop(
    "Missing sample summary file(s):\n",
    paste(missing_paths, collapse = "\n")
  )
}

summary_rows <- Map(
  function(path, configured_sample_id) {
    sample_row <- read_tsv(path, show_col_types = FALSE)

    if (nrow(sample_row) == 0) {
      stop("Sample summary file is empty: ", path)
    }

    # Older sample-summary files do not always contain a sample_id column.
    # Prefer an existing sample/sample_id value, but fall back to the sample
    # identifier supplied by Snakemake and used to construct this file path.
    if (!"sample_id" %in% names(sample_row)) {
      if ("sample" %in% names(sample_row)) {
        sample_row <- sample_row |>
          rename(sample_id = sample)
      } else {
        sample_row$sample_id <- configured_sample_id
      }
    }

    sample_row$sample_id <- as.character(sample_row$sample_id)
    missing_sample_id <- is.na(sample_row$sample_id) |
      !nzchar(trimws(sample_row$sample_id))
    sample_row$sample_id[missing_sample_id] <- configured_sample_id

    sample_row
  },
  sample_summary_paths,
  sample_ids
)

summary <- bind_rows(summary_rows)

# Normalize segment-count column names across sample-summary versions.
# The current sample-summary generator uses `segments_pass`, while some older
# run-summary tables used `pass_segments`.
if (!"pass_segments" %in% names(summary)) {
  if ("segments_pass" %in% names(summary)) {
    summary <- summary |>
      rename(pass_segments = segments_pass)
  } else if ("coverage_segments" %in% names(summary)) {
    summary <- summary |>
      mutate(pass_segments = coverage_segments)
  } else if ("segments_detected" %in% names(summary)) {
    summary <- summary |>
      mutate(pass_segments = segments_detected)
  } else {
    stop(
      "The sample summary files do not contain any recognized passing-segment ",
      "column (expected one of: pass_segments, segments_pass, ",
      "coverage_segments, segments_detected)."
    )
  }
}

summary <- summary |>
  mutate(
    pass_segments = suppressWarnings(as.numeric(pass_segments)),
    pass_segments = coalesce(pass_segments, 0)
  )

# Retain a normalized detected-segment count when available.
if (!"coverage_segments" %in% names(summary)) {
  if ("segments_detected" %in% names(summary)) {
    summary <- summary |>
      mutate(coverage_segments = suppressWarnings(as.numeric(segments_detected)))
  } else {
    summary$coverage_segments <- summary$pass_segments
  }
}

if (!"genome_status" %in% names(summary)) {
  summary <- summary |>
    mutate(
      genome_status = case_when(
        pass_segments == 8 ~ "Complete",
        pass_segments >= 6 ~ "Near-complete",
        pass_segments >= 1 ~ "Partial",
        TRUE ~ "Failed"
      )
    )
}

if (!"review_required" %in% names(summary)) {
  if ("review_flags" %in% names(summary)) {
    normalized_flags <- toupper(trimws(as.character(summary$review_flags)))
    flag_review <- !is.na(normalized_flags) &
      nzchar(normalized_flags) &
      !normalized_flags %in% c("NONE", "NA", "N/A", "CLEAR", "OK")
    summary$review_required <- flag_review | summary$pass_segments < 8
  } else {
    summary$review_required <- summary$pass_segments < 8
  }
} else if (is.logical(summary$review_required)) {
  summary$review_required <- coalesce(summary$review_required, FALSE)
} else {
  summary$review_required <-
    tolower(trimws(as.character(summary$review_required))) %in%
    c("true", "t", "1", "yes", "y")
}

if (!"review_reason" %in% names(summary)) {
  if ("review_flags" %in% names(summary)) {
    flags <- trimws(as.character(summary$review_flags))
    flags[toupper(flags) %in% c("NONE", "NA", "N/A", "CLEAR", "OK", "")] <- NA_character_
    summary$review_reason <- ifelse(
      summary$review_required,
      ifelse(
        is.na(flags),
        paste0("Genome status: ", summary$genome_status),
        gsub(";", "; ", flags, fixed = TRUE)
      ),
      NA_character_
    )
  } else {
    summary$review_reason <- ifelse(
      summary$review_required,
      paste0("Genome status: ", summary$genome_status),
      NA_character_
    )
  }
}

# Derive compact HA/NA calls from contig names when explicit calls are absent.
extract_subtype <- function(values, prefix) {
  values <- as.character(values)
  pattern <- paste0(".*(", prefix, "[0-9]+).*")
  matched <- grepl(pattern, values, ignore.case = TRUE)
  result <- rep(NA_character_, length(values))
  result[matched] <- toupper(sub(pattern, "\\1", values[matched], ignore.case = TRUE))
  result
}

if (!"ha_call" %in% names(summary) && "ha_contig" %in% names(summary)) {
  summary$ha_call <- extract_subtype(summary$ha_contig, "H")
}

if (!"na_call" %in% names(summary) && "na_contig" %in% names(summary)) {
  summary$na_call <- extract_subtype(summary$na_contig, "N")
}

if (!"potential_subtype" %in% names(summary)) {
  ha_value <- if ("ha_call" %in% names(summary)) as.character(summary$ha_call) else NA_character_
  na_value <- if ("na_call" %in% names(summary)) as.character(summary$na_call) else NA_character_
  summary$potential_subtype <- case_when(
    !is.na(ha_value) & nzchar(ha_value) & !is.na(na_value) & nzchar(na_value) ~ paste0(ha_value, na_value),
    !is.na(ha_value) & nzchar(ha_value) ~ ha_value,
    !is.na(na_value) & nzchar(na_value) ~ na_value,
    TRUE ~ "Undetermined"
  )
}

if (!"subtype_min_median_depth" %in% names(summary)) {
  ha_depth <- if ("ha_median_depth" %in% names(summary)) suppressWarnings(as.numeric(summary$ha_median_depth)) else rep(NA_real_, nrow(summary))
  na_depth <- if ("na_median_depth" %in% names(summary)) suppressWarnings(as.numeric(summary$na_median_depth)) else rep(NA_real_, nrow(summary))
  summary$subtype_min_median_depth <- ifelse(
    is.na(ha_depth) & is.na(na_depth),
    NA_real_,
    pmin(ha_depth, na_depth, na.rm = TRUE)
  )
}

column_defaults <- list(
  host = "Unknown",
  potential_subtype = "Undetermined",
  ha_call = NA_character_,
  na_call = NA_character_,
  subtype_min_median_depth = NA_real_,
  aa_changes = NA_real_,
  synonymous_changes = NA_real_,
  nonsynonymous_changes = NA_real_
)

for (column_name in names(column_defaults)) {
  if (!column_name %in% names(summary)) {
    summary[[column_name]] <- column_defaults[[column_name]]
  }
}

# Normalize missing subtype calls for display and export. The influenza A
# segment name "NA" is retained; only absent subtype calls become "Unknown".
normalize_subtype_call <- function(values) {
  values <- trimws(as.character(values))
  missing_call <- is.na(values) |
    !nzchar(values) |
    toupper(values) %in% c("NA", "N/A", "NONE", "NO_HIT", "UNDETERMINED")
  values[missing_call] <- "Unknown"
  values
}

summary <- summary |>
  mutate(
    ha_call = normalize_subtype_call(ha_call),
    na_call = normalize_subtype_call(na_call),
    potential_subtype = normalize_subtype_call(potential_subtype)
  ) |>
  arrange(sample_id)

review <- summary |>
  filter(review_required)

write_tsv(summary, params$run_summary_tsv)
write_tsv(review, params$review_tsv)

host_values <- if ("host" %in% names(summary)) {
  unique(na.omit(summary$host))
} else {
  character(0)
}

metadata <- tibble(
  Metric = c("Run date", "Host", "Pipeline version"),
  Value = c(
    format(Sys.time(), "%Y-%m-%d %H:%M:%S"),
    if (length(host_values) > 0) paste(host_values, collapse = ", ") else "Not available",
    "1.0.0"
  )
)
```

```{=html}
<div class="report-header">
  <div class="report-brand-mark" aria-hidden="true">
    <span></span>
    <span></span>
    <span></span>
  </div>

  <div class="report-title-group">
    <div class="report-title">
      WINGS — Wild-bird Influenza Genomics and Surveillance
    </div>
    <div class="report-subtitle">
      Run Summary Report
    </div>
  </div>

  <div class="report-local-badge">
    <span class="report-local-dot"></span>
    Local analysis
  </div>
</div>
```

## Overview

```{r}
total_samples <- nrow(summary)

complete_n <- sum(summary$pass_segments == 8, na.rm = TRUE)
near_complete_n <- sum(
  summary$pass_segments >= 6 & summary$pass_segments < 8,
  na.rm = TRUE
)
partial_n <- sum(
  summary$pass_segments >= 1 & summary$pass_segments < 6,
  na.rm = TRUE
)
failed_n <- sum(summary$pass_segments == 0, na.rm = TRUE)

overview <- tibble(
  Metric = c("Samples", "Complete", "Near-complete", "Partial", "Failed"),
  Count = c(
    total_samples,
    complete_n,
    near_complete_n,
    partial_n,
    failed_n
  ),
  Percent = c(
    1,
    complete_n / max(total_samples, 1),
    near_complete_n / max(total_samples, 1),
    partial_n / max(total_samples, 1),
    failed_n / max(total_samples, 1)
  )
)

overview |>
  gt(rowname_col = "Metric") |>
  tab_header(
    title = "Run Overview",
    subtitle = "QC-qualified genome completeness across all samples"
  ) |>
  fmt_percent(columns = Percent, decimals = 0) |>
  cols_label(
    Count = "Count",
    Percent = "Percent"
  ) |>
  tab_style(
    style = cell_fill(color = "#F2F2F2"),
    locations = cells_body(rows = Metric == "Samples")
  ) |>
  tab_style(
    style = cell_fill(color = "#E7F4EA"),
    locations = cells_body(rows = Metric == "Complete")
  ) |>
  tab_style(
    style = cell_fill(color = "#FFF3CD"),
    locations = cells_body(rows = Metric == "Near-complete")
  ) |>
  tab_style(
    style = cell_fill(color = "#FCE8DC"),
    locations = cells_body(rows = Metric == "Partial")
  ) |>
  tab_style(
    style = cell_fill(color = "#F4D6DC"),
    locations = cells_body(rows = Metric == "Failed")
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#8C1D40"),
      cell_text(color = "white")
    ),
    locations = cells_title(groups = "title")
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "#8C1D40"),
      cell_text(color = "white")
    ),
    locations = cells_title(groups = "subtitle")
  ) |>
  tab_options(
    table.width = pct(100),
    table.font.size = px(14),
    heading.title.font.size = px(20),
    heading.subtitle.font.size = px(14),
    column_labels.font.weight = "normal",
    data_row.padding = px(8)
  )
```

```{r}
metadata |>
  gt() |>
  tab_options(
    table.width = pct(60),
    table.font.size = px(14),
    data_row.padding = px(8)
  )
```

## ⚠ Samples Requiring Immediate Review

```{r}
n_review <- nrow(review)

if (n_review == 0) {
  div(
    class = "alert alert-success",
    h4("✓ No samples require manual review."),
    p("All samples passed the configured QC thresholds.")
  )
} else {
  div(
    class = "alert alert-warning",
    h4(sprintf("%d sample(s) require manual review", n_review)),
    p("These samples failed one or more quality-control criteria. Review these before downstream analyses.")
  )

  review |>
    mutate(
      sample_id = sprintf(
        "[%s](../%s/summary/%s.sample_summary.html)",
        sample_id,
        sample_id,
        sample_id
      )
    ) |>
    select(
      sample_id,
      host,
      potential_subtype,
      genome_status,
      review_reason
    ) |>
    gt() |>
    fmt_markdown(columns = sample_id) |>
    cols_label(sample_id = "Sample") |>
    tab_options(
      table.width = pct(100),
      table.font.size = px(13),
      heading.background.color = "#8C1D40",
      heading.title.font.size = px(18)
    )
}
```

## Sample Summary

```{r}
sample_table <-
  summary %>%
  mutate(
    sample_id = sprintf(
      "[%s](../%s/summary/%s.sample_summary.html)",
      sample_id,
      sample_id,
      sample_id
    ),
    QC = genome_status,
    Review = if_else(
      review_required,
      "REVIEW",
      "PASS"
    )
  ) %>%
  select(
    sample_id,
    QC,
    Review,
    host,
    potential_subtype,
    ha_call,
    na_call,
    pass_segments,
    subtype_min_median_depth,
    aa_changes,
    synonymous_changes,
    nonsynonymous_changes
  )

sample_table |>
  gt() |>
  tab_header(
    title = "Sample Summary"
  ) |>
  cols_label(
    sample_id = "Sample",
    QC = "QC Status",
    Review = "Review",
    host = "Host",
    potential_subtype = "Predicted Subtype",
    ha_call = "HA",
    na_call = "NA",
    pass_segments = "Passing Segments",
    subtype_min_median_depth = "Min Median Depth",
    aa_changes = "AA Changes",
    synonymous_changes = "Synonymous",
    nonsynonymous_changes = "Nonsynonymous"
  ) |>
  fmt_markdown(columns = sample_id) |>
  tab_options(
    table.width = pct(100),
    table.font.size = px(13),
    heading.background.color = "#8C1D40",
    heading.title.font.size = px(18)
  )
```

## QC-qualified Genome Completeness

Each influenza segment passes WINGS segment QC when median depth is at least **`r format(coverage_threshold, trim = TRUE)`x**; at least **`r format(100 * coverage_breadth_threshold, trim = TRUE)`%** of reference positions have depth of at least **`r format(coverage_threshold, trim = TRUE)`x**; consensus length is at least the configured segment-specific minimum; and the consensus N fraction is at most **`r format(100 * max_n_fraction, trim = TRUE)`%**. A consensus longer than the configured upper length guide is retained as **PASS** but receives a length warning for review. Genome status is based on the number of the eight influenza A segments that pass the hard QC criteria:

- **Complete:** 8 of 8 segments pass segment QC
- **Near-complete:** 6-7 of 8 segments pass segment QC
- **Partial:** 1-5 of 8 segments pass segment QC
- **Failed:** 0 of 8 segments pass segment QC

These categories describe **QC-qualified genome recovery** and should not be interpreted as a substitute for downstream annotation, biological review, or other assay-specific quality checks.

```{r}
asu_palette <- c(
  Complete = "#8C1D40",
  "Near-complete" = "#A95670FF",
  Partial = "#C58E9FFF",
  Failed = "#DFC9CFFF"
)

genome_status <-
  summary %>%
  count(genome_status) %>%
  mutate(
    genome_status = factor(
      genome_status,
      levels = c("Complete", "Near-complete", "Partial", "Failed")
    )
  )

p <- ggplot(
  genome_status,
  aes(
    genome_status,
    n,
    fill = genome_status
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(label = n),
    vjust = -0.3,
    size = 5
  ) +
  scale_fill_manual(values = asu_palette) +
  labs(
    x = "Genome Status",
    y = "Samples"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none"
  )

p
```

## HA Subtype Distribution

```{r}
ha_counts <- summary %>%
  count(ha_call)

p <- ggplot(
  ha_counts,
  aes(
    x = reorder(ha_call, n),
    y = n
  )
) +
  geom_col(
    fill = "#8C1D40",
    width = 0.7
  ) +
  coord_flip() +
  labs(
    x = "HA Subtype",
    y = "Samples"
  ) +
  theme_minimal(base_size = 14)

p
```

## NA Subtype Distribution

```{r}
na_counts <- summary %>%
  count(na_call)

p <- ggplot(
  na_counts,
  aes(
    x = reorder(na_call, n),
    y = n
  )
) +
  geom_col(
    fill = "#FFC627",
    width = 0.7
  ) +
  coord_flip() +
  labs(
    x = "NA Subtype",
    y = "Samples"
  ) +
  theme_minimal(base_size = 14)

p
```

## Host Distribution

```{r}
host_counts <-
  summary %>%
  count(host)

p <- ggplot(
  host_counts,
  aes(
    reorder(host, n),
    n
  )
) +
  geom_col(
    fill = "#8C1D40",
    width = 0.7
  ) +
  coord_flip() +
  labs(
    x = "Host",
    y = "Samples"
  ) +
  theme_minimal(base_size = 14)

p
```

## Average Coverage

```{r}
p <- ggplot(
  summary,
  aes(subtype_min_median_depth)
) +
  geom_histogram(
    bins = 30,
    fill = "#8C1D40",
    color = "white"
  ) +
  labs(
    x = "Mean depth",
    y = "Samples"
  ) +
  theme_minimal(base_size = 14)

p
```

## Sequencing Depth Distribution

```{r}
p <- ggplot(
  summary,
  aes(pass_segments)
) +
  geom_histogram(
    binwidth = 1,
    fill = "#FFC627",
    color = "white"
  ) +
  scale_x_continuous(breaks = 0:8) +
  labs(
    x = "Passing genome segments",
    y = "Samples"
  ) +
  theme_minimal(base_size = 14)

p
```

## Publication-quality Summary Figure

::: {.center-figure}

```{r}
figure_path <- "figures/run_summary.png"

if (file.exists(figure_path)) {
  knitr::include_graphics(figure_path)
} else {
  cat("Publication-quality summary figure was not generated for this run.")
}
```

:::
